---
title: "Lab_5"
author: "Terry Tian"
date: "2025-02-05"
output: html_document
---

```{r include=FALSE}
# install.packages("kableExtra")

library(data.table)
library(tidyverse)
library(leaflet)
library(reticulate)
library(kableExtra)
library(dplyr)
library(dtplyr)
library(mgcv)

```


Download MET and Stations Datsets

```{r message = FALSE}

# Download the data
stations <- fread("ftp://ftp.ncdc.noaa.gov/pub/data/noaa/isd-history.csv")
stations[, USAF := as.integer(USAF)]

# Dealing with NAs and 999999
stations[, USAF   := fifelse(USAF == 999999, NA_integer_, USAF)]
stations[, CTRY   := fifelse(CTRY == "", NA_character_, CTRY)]
stations[, STATE  := fifelse(STATE == "", NA_character_, STATE)]

# Selecting the three relevant columns, and keeping unique records
stations <- unique(stations[, list(USAF, CTRY, STATE)])

# Dropping NAs
stations <- stations[!is.na(USAF)]

# Removing duplicates
stations[, n := 1:.N, by = .(USAF)]
stations <- stations[n == 1,][, n := NULL]

# Read in the met data
download.file(
  "https://raw.githubusercontent.com/JSC370/JSC370-2025/main/data/met/met_all.gz",
  destfile = "met_all.gz",
  method   = "curl",
  timeout  = 60
  )

met <- data.table::fread("met_all.gz")

```

Merge the Data
```{r message=FALSE}

# Using tidyverse package
met <- left_join(
  x = met,
  y = stations,
  by = c( "USAFID" = "USAF")
)


# Or using merge
# met <- merge(
#   x=met,
#   y=stations,
#   all.x = TRUE,
#   all.y = FALSE,
#   by.x = "USAFID",
#   by.y = "USAF"
# )
```


## Question 1 - Find stations with media value of temp, wind speed and pressure


```{r}

medians <- met[, .(
  temp_50 = quantile(temp, probs = 0.5, na.rm=TRUE),
  wind_sp_50 = quantile(wind.sp, probs = 0.5, na.rm=TRUE),
  pressure_50 = quantile(atm.press, probs = 0.5, na.rm=TRUE)
)]

medians

```

Group by USAFID and State

```{r message = FALSE, output=FALSE}

station_med <- met[, .(
  temp = quantile(temp, probs = 0.5, na.rm=TRUE),
  wind.sp = quantile(wind.sp, probs = 0.5, na.rm=TRUE),
  atm.press = quantile(atm.press, probs = 0.5, na.rm=TRUE),
  lat = mean(lat, na.rm=TRUE),
  lon = mean(lon, na.rm=TRUE)
), by = .(USAFID, STATE)]


head(station_med)

```

Compute absolute distance from median and find station equal to median
```{r message = FALSE, results = 'hide'}

# Median temp stations
station_med[, temp_dist := abs(temp - medians$temp_50)]
median_temp_station <- station_med[temp_dist == 0]
median_temp_station

# Wind Speed
station_med[, wind.sp_dist := abs(wind.sp - medians$wind_sp_50)]
median_wind_station <- station_med[wind.sp_dist == 0]
median_wind_station

# Atm Pressure
station_med[, pressure_dist := abs(atm.press - medians$pressure_50)]
median_pressure_station <- station_med[pressure_dist == 0]
median_pressure_station

```

## Quetion 2 -Compute closest by Euclidean distance

```{r results = 'hide'}

station_med[, temp_50 := quantile(temp, probs=0.5, na.rm=TRUE), by=STATE]
station_med[, wind.sp_50 := quantile(wind.sp, probs=0.5, na.rm=TRUE),  by=STATE]

# Get the euclidean distance
station_med[, eudist := sqrt(
  (temp - temp_50)^2 + (wind.sp - wind.sp_50)^2
)]

# Choose the most representative, the median, station per state with the lowest euclidean distance
id_station <- station_med[, .SD[which.min(eudist)], by=STATE]

id_station <- merge(
  x = id_station,
  y = stations,
  by.x = "USAFID", 
  by.y = "USAF",
  all.x = TRUE,
  all.y = FALSE
)

id_station

```


## Question 3 - Get closest station to midpoint
```{r results = 'hide'}

# Get mid point of state

midpoint <- met[, .(
  lon_50 = quantile(lon, probs = 0.5, na.rm=TRUE),
  lat_50 = quantile(lat, probs = 0.5, na.rm=TRUE)
), by = STATE]


mid <- merge(
  x = met,
  y = midpoint,
  by = "STATE"
)

# Calculate euclidean dist for lon and lat
mid[, mid_euclid := sqrt(
  (lon - lon_50)^2 + (lat - lat_50)^2
)]

# Find closest station to mid-point
closest <- mid[, .SD[which.min(mid_euclid)], by = STATE]


```

Visualize Center and Closest Station
```{r echo=FALSE}

leaflet() %>% 
  addProviderTiles('CartoDB.Positron') %>%
  addCircles(
    data = closest,
    lat = ~lat, lng = ~lon, popup = "geographic mid station",
    opacity = 1, fillOpacity = 1, radius = 400, color = 'blue'
  )%>% 
  addCircles(
    data = id_station,
    lat = ~lat, lng = ~lon, popup = "eudist mid station",
    opacity = 1, fillOpacity = 1, radius = 400, color = 'purple'
  )

```


## Question 4 - Generate summary table

Start by computing the states’ average temperature. Use that measurement to classify them according to the following criteria:

```{r }

met <- met %>% mutate(
  avg_temp_lvl = cut(temp,
                        breaks = c(-Inf, 10, 20, Inf),
                        labels = c("low", "med", "high"),
                        right = FALSE))

# Create summary table
summary <- met %>% 
  group_by(avg_temp_lvl) %>% 
  summarise(
    n_states = n_distinct(STATE),
    avg_temp = mean(temp, na.rm=T),
    avg_wind.sp = mean(wind.sp, na.rm=T),
    avg_pressure = mean(atm.press, na.rm=T)
  )


kable(summary, booktabs = TRUE) %>% 
  kable_styling(font_size = 10) %>% 
  kable_paper("hover", full_width = F)

```

## Question 5 - Advanced Regression


```{r }

station_med_lt <- lazy_dt(station_med)
station_med_lt <- station_med_lt |> filter(between(atm.press, 1000, 1020)) |> collect()


ggplot(station_med_lt, aes(x=atm.press, y=temp)) +
  geom_point() +
  geom_smooth(method="lm", col="cyan") +
  geom_smooth(method="gam", col="blue")

```

```{r }

lm_mod <- lm(temp ~ atm.press, data=station_med_lt)
summary(lm_mod)

```

GAM
```{r}

gam_mod <- gam(temp ~ s(atm.press, bs="cr", k=20), data=station_med_lt)
summary(gam_mod)
plot(gam_mod)


```



